Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-12-08 rnni_convexity.tex
to induction step of lemma caterpillar_dist

The above observation suggests that there is a caterpillar path of length $d(T,R)$ between any two caterpillar trees $T,R$ in $\rnni$.
Indeed, we are going to prove within this section that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex}).
Note that this is not true for the $\nni$ space, which is clear from the example in Figure~\ref{fig:NNI_vs_RNNI}.

The following lemma regards caterpillar trees with distance $\frac{(n-1)(n-2)}{2}$ and will be needed for proving the convexity of the set of caterpillar trees in $\rnni$.
Notice that $\frac{(n-1)(n-2)}{2}$ is the upper bound for the diameter of $\rnni$ by Lemma~\ref{lemma:diameter_bound}.

@@

\begin{proof}
Let us explain that if $d(T,R) = \frac{(n-1)(n-2)}{2}$, we can follow $d_c(T,R) = \frac{(n-1)(n-2)}{2}$ by using $\csort$:
In each step $i=0, \ldots, n-3$ of $\csort$ at most $n-2-i$ pairs of taxa swap positions.
This is due to the fact that in worst case the taxon that is considered in step $k$ is moved from the cherry of the tree up to position $n-k$.
It follows that the maximum length of a path computed by $\csort$ is $\sum\limits_{i=0}^{n-3} n-2-i = \frac{(n-1)(n-2)}{2}$.
Since there cannot be a caterpillar path shorter than $d(T,R) = \frac{(n-1)(n-2)}{2}$, it follows $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.

For proving the other direction of the statement, assume that $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
We prove $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$ of $T$ and $R$:

For the induction basis we consider the $\rnni$ graph $n=3$ taxa which consists of three trees only.
Since every of these trees is a caterpillar tree that is connected with both other trees by an edge, it is $d(T,R) = d_c(T,R) = 1$ for all pairs of trees $T,R$ on $n=3$ taxa.

For the induction step we assume without loss of generality that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$ and $R$ is a caterpillar tree with distance $\frac{n(n-1)}{2}$ to $T$.
It follows that when applying $\csort$ onto $T,R$ the number of taxon swaps is maximum in each step, which implies that taxon $n+1$ is part of the cherry of $R$.
Let us now consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
$\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same $\nni$ moves as $p$, except for the $n-1$ ones on $p$ that involve taxon $n+1$.
Hence, $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
So we can apply the induction hypothesis on $T{\big|}_n$ and $R{\big|}_n$ and know that $d(T{\big|}_n,R{\big|}_n) = \frac{(n-1)(n-2)}{2}$.
With Lemma~\ref{lemma:distance_delete_taxon} we also know that it is $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| = d(T,R) - (n-1)$, where the last equality follows from our observation above.

>>>>

These observations motivate the following general result.
Every pair of trees $T, R$ in $\rnni$ are connected by a caterpillar path of length $d(T,R)$, that is, the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex}).
Note that this is not true in the $\nni$ space (Figure~\ref{fig:NNI_vs_RNNI}).

We need the following lemma.
Recall that $\frac{(n-1)(n-2)}{2}$ is the upper bound of the diameter of $\rnni$ by Lemma~\ref{lemma:diameter_bound}.

@@

\begin{proof}
First note that $d(T,R) = \frac{(n-1)(n-2)}{2}$ implies $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.Indeed, in each step $i=0, \ldots, n-3$ of $\csort$ at most $n-2-i$ pairs of taxa swap positions, as in the worst case the taxon considered in step $k$ is moved from the cherry of the tree up to the position $n-k$.
It follows that the maximum length of a path computed by $\csort$ is $\sum\limits_{i=0}^{n-3} n-2-i = \frac{(n-1)(n-2)}{2}$.

For proving the other direction of the statement, assume that $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
We prove that $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$. The induction basis for $n=3$ taxa is trivial as the $\rnni$ graph is a triangle on three caterpillar trees.

For the induction step we assume without loss of generality that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$ and $R$ is a caterpillar tree at distance $\frac{n(n-1)}{2}$ to $T$.
It follows that when applying $\csort$ onto $T,R$ the number of taxon swaps is maximum in each step, which implies that taxon $n+1$ is part of the cherry of $R$.
Let us now consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
$\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same $\nni$ moves as $p$, except for the $n-1$ ones on $p$ that involve taxon $n+1$.
Hence, $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
So we can apply the induction hypothesis on $T{\big|}_n$ and $R{\big|}_n$ and know that $d(T{\big|}_n,R{\big|}_n) = \frac{(n-1)(n-2)}{2}$.
With Lemma~\ref{lemma:distance_delete_taxon} we also know that it is $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| = d(T,R) - (n-1)$, where the last equality follows from our observation above.
